###Loading and preprocessing the data###    
```{r load, echo = TRUE}
data <- read.csv ("./activity.csv")
# convert the variable 'date' from 'Factor' to 'Date'
data$date <- as.Date(data$date)
```

``` {r libraries, echo = TRUE}
library (dplyr)
library (ggplot2)
```

###What is the mean number of steps taken by day?###    
 
1. Calculate the total number of steps taken per day  
```{r tot.per.day, echo = TRUE}
steps_day <- data %>% group_by (date) %>% summarize (tot = sum (steps, na.rm = TRUE))
```
  
2. Make a histogram of the total number of steps taken each day      
```{r hist.tot, echo = TRUE}
ggplot (data = steps_day, aes (tot)) + 
        geom_histogram (col = "grey") +
        labs (x = "Number of steps taken each day") +
        labs (y = "Frequency (number of days)") +
        labs (title = "Histogram of total number of steps taken each day")
```

3. Calculate and report the mean and median of the total number of steps taken each day    
```{r mean, echo = TRUE}
steps_day_mean <- mean (steps_day$tot)
steps_day_mean
```

```{r median, echo = TRUE}
steps_day_median <- median (steps_day$tot)
steps_day_median
```
    
    
###What is the average daily activity pattern?###  
  
1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days  
```{r plot.5min, echo = TRUE}
mean_int <- data %>% group_by (interval) %>% summarize (mean = mean (steps, na.rm = TRUE))
ggplot (data = mean_int, aes( interval, mean)) + 
        geom_line() + 
        labs (x = "5 minute interval") +
        labs (y = "Average number of steps taken") +
        labs (title = "Time series plot of the 5-minute interval \n and the average number of steps taken, averaged across all days ")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r max.steps, echo = TRUE}
mean_int %>% filter (mean == max (mean_int$mean))
```
  
  
###Imputing missing values###  

1. Calculate and report the total number of missing values in the dataset  
```{r na.values, echo = TRUE}
sum (is.na (data$steps))
mean (is.na (data$steps)) * 100
```


2. Devise a strategy for filling in all the missing values.   
The strategy for replacing a missing value is to use the mean value of steps taken in the related 5-minute interval, averaged across all days  

3. Create a dataset that is equal to the original dataset, but with the missed data filled in  
```{r data.fill, echo = TRUE}
data_fill <- data
for (i in 1 : nrow (data_fill))
        {
        if (is.na (data_fill$steps[i]))
                {
                data_fill$steps[i] <- floor (subset (mean_int, interval == data_fill$interval[i])$mean)
                }
        }
```
  
4. Make a histogram of the total number of steps taken each day and calculate and report the **mean** and **median** total number of steps taken per day.   
  
4a. Histogram of the total number of steps taken each day  
```{r hist.tot.fill, echo = TRUE, fig.width = 7}
steps_day_fill <- data_fill %>% group_by (date) %>% summarize (tot = sum (steps))
ggplot (data = steps_day_fill, aes (tot)) + 
        geom_histogram (color = "black") +
        labs (x = "Number of steps taken each day") +
        labs (y = "Frequency (number of days)") +
        labs (title = "Histogram of total number of steps taken each day. \n (A missing value has been replaced by the mean value \n of the related 5-minute interval, averaged across all days)")
```

4b. Calculate and report mean and median total number of steps taken per day  
```{r mean.fill, echo = TRUE}
steps_day_fill_mean <- mean (steps_day_fill$tot)
steps_day_fill_mean
```

```{r median.fill, echo = TRUE}
steps_day_fill_median <- median (steps_day_fill$tot)
steps_day_fill_median
```

4c. Do these values differ from the estimates from the first part of the assignement?  
```{r compare1, echo = TRUE, fig.width = 8}
# merge the original datasets  with the imputed dataset
steps_day_all <- rbind (steps_day, steps_day_fill)
# add a new variable indicating wich dataset the data come from
steps_day_all <- mutate (steps_day_all, flag = factor (rep (c ("Non-imputed NAs", "Imputed NAs"), c (nrow (steps_day), nrow (steps_day_fill)))))
# plot the 'global' dataset as histogram
ggplot (data = steps_day_all, aes (tot)) + 
        geom_histogram () +
        facet_grid (. ~ flag) +
        labs (x = "Number of steps taken each day") +
        labs (y = "Frequency (number of days)")
```

4d. What is the impact of imputing missing data on the estimates of the total daily number of steps? 

1. The number of days with the lowest (zero) number of steps has decreased  
2. The number of days with the highest number of steps has increased 
3. The mean and the median of the total daily number of steps have been modified as follows:  
```{r compare2, echo = TRUE}
compare <- steps_day_all %>% group_by (flag) %>% summarise_each (funs (mean (., na.rm = TRUE), median (., na.rm = TRUE)), tot) 
compare
```
  
  
###Are there differences in activity patterns between weekdays and weekends?###  
(Use the dataset with the filled-in missing values)  

1. Create a new factor variable in the dataset with 2 levels, `weekday` and `weekend`
```{r weekday.weekend, echo = TRUE}
data_fill <- data_fill %>% mutate (day_type = "weekday") 
for (i in 1: nrow (data_fill))
        {
        if (as.POSIXlt (as.Date (data_fill$date[i]))$wday %in% c(6, 0))
                {
                data_fill$day_type[i] <- c("weekend")
                }
}
data_fill$day_type <- factor (data_fill$day_type)
```

2. Make a panel plot containing a time series of the 5-minutes interval and the average numbers of steps taken, averaged across all weekdays and all weekend days.
```{r plot.week, fig.width = 9, echo = TRUE}
mean_day_int <- data_fill %>% group_by (day_type, interval) %>% summarize (mean = mean (steps))
ggplot (data = mean_day_int, aes (interval, mean)) +
        geom_line () +
        facet_grid (. ~ day_type) +
        labs (x = "5 minute interval") +
        labs (y = "Average number of steps taken") +
        labs (title = "Time series plot of the 5-minute interval \n and the average number of steps taken,\n averaged across all weekdays and all weekends ")
```

